---
title: "Biomass_core"
author:
  - Jean Marchal (<jean.d.marchal@gmail.com>)
  - Ceres Barros (<cbarros@mail.ubc.ca>)
  - Eliot McIntire (<eliot.mcintire@canada.ca>)
  - Yong Luo
  - Alex M. Chubaty (<achubaty@for-cast.ca>)
date: "11 May 2021"
output:
  html_document:
    df_print: paged
    keep_md: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning = FALSE, results = "hold")
```

[![Gitter](https://badges.gitter.im/PredictiveEcology/LandR_Biomass.svg)](https://gitter.im/PredictiveEcology/LandR_Biomass?utm_source=badge&utm_medium=badge&utm_campaign=pr-badge)

**This documentation is not complete. If there are discrepancies, omissions, or it is unclear, this is expected**

# Overview

A biomass succession model derived and modified from LANDIS-II Biomass Succession v3.2.1 (hereafter, LBSE).
Biomass_core simulates forest succession in a spatially explicit manner (per pixel) by reproducing the population dynamics (growth and mortality), dispersal and competition for light resources of tree species. 
Like in LBSE, dynamics are simulated in terms of their biomass per cohort (a combination of species and age), following a `biomass ~ age` curve  that is influenced by species-specific parameters such as growth, mortality, maximum biomass, and species traits such as longevity and tolerance to shade (see [LANDIS-II Biomass Succession v3.2 User Guide](https://github.com/LANDIS-II-Foundation/Extension-Biomass-Succession/blob/master/docs/LANDIS-II%20Biomass%20Succession%20v3.2%20User%20Guide.docx) and [Scheller and Mladenoff (2004)](https://pdfs.semanticscholar.org/4d38/d0be6b292eccd444af399775d37a757d1967.pdf).
One of three dispersal algorithms are available: 'no dispersal', 'universal dispersal'(every species can disperse across the entire landscape) and 'ward dispersal' (species disperse according to a dispersal kernel, with mean and maximum distances defaulting to LBSE trait table values). Ward dispersal is used by default.
The biggest differences between Biomass_core and LBSE lie on how the model is parametrized and initialized, how different processes have been compartmentalized in order for higher flexibility, the use of hashing, the concurrent calculation of cohort growth and mortality (instead of following a particular species order) and a minor change to the calculation of total biomass (*sumB*) when assigning biomass (*B*) to new cohorts - in LBSE, depending on whether the succession time step was = 1 or not, other new cohorts where included/excluded from *sumB*.

## Inputs and parametrisation

`Biomass_core` is capable of running on dummy datasets from which it estimates parameters linked to vegetation growth and seed germination (such as the maximum biomass per species, per pixel, and the probability of seed germination - *i.e.*, species establishment probability not due to resprouting), but also builds and initializes forest communities (based on biomass, age, species composition, land cover and ecological zones like ecodistricts.

Ideally, however, the user should supply realistic versions of these data and the essential initialization objects that `Biomass_core` requires to run.

```{r moduleInputs, eval = TRUE, echo = FALSE}
df_inputs <- SpaDES.core::moduleInputs("Biomass_core", "..")
knitr::kable(df_inputs)
```

Below is a list of input objects that the user (or another module) should supply to `Biomass_core` in order for it to run on a realistic setting (this is not, however, an exhaustive list of inputs). The only input that **must** be provided to `Biomass_core` is `studyArea`

Type                       | Required inputs    | Description
-------------------------- | -----------------  | -------------------------------------------------------------------------------
Spatial layers             |                    |
| `biomassMap`       | map of total biomass in the study area
| `ecoregionMap`     | map of ecological zones ('ecolocations') in the study area
| `rasterToMatch`    | the map of the study area.
| `studyArea`        | a spatial polygon of the study area. MUST be provided.
Trait and parameter tables |                    |
| `ecoregion`        | a table listing values in `ecoregionMap` and their state (active or not)
| `minRelativeB`     | table defining the cut points to classify stand shadiness
| `species`          | a table of species traits (based on LANDIS-II species traits for [v3.2.1] (https://github.com/LANDIS-II-Foundation/Extensions-Succession-Archive/blob/master/biomass-succession-archive/trunk/tests/v6.0-2.0/species.txt))
| `speciesEcoregion` | a table of species traits by for ecological zone in ecoregionMap
| `sppEquiv`         | a table of with species names and their correspondences across multiple datasets (see ?LandR::sppEquivalencies_CA)      
| `sppColorVect`     | a table of species traits by for ecological zone in ecoregionMap          
| `sufficientLight`  | table defining species responses to shade according to shade tolerance levels
Cohort-simulation related  |                    |
| `cohortData`       | table with cohort information by pixelGroup 
| `pixelGroupMap`    | map of pixelGroups

For the beginner user, we suggest running `Biomass_core` without supplying any inputs and inspecting the above mentioned objects to understand their structure and format.
The user can later either feed these objects via `simInit`, or make a module that makes them and provides necessary inputs to `Biomass_core` (see e.g. [Boreal_Biomass_coreDataPrep](https://github.com/PredictiveEcology/Boreal_Biomass_coreDataPrep))

Besides the above mentioned inputs, `Biomass_core` uses several other parameters, which can be changed by the user if need be. Below is a list of the most useful ones, but please check module metadata for full list and descriptions.

Type              | Required inputs        | Description
----------------- | ---------------------- | -------------------------------------------------------------------------------
Plotting & saving |                        |
| `.plots`               | activates/deactivates plotting and defines type fo plotting (see `?Plots`)
| `.plotInitialTime`     | defines when plotting starts 
| `.plotInterval`        | defines plotting frequency
| `.plotMaps`            | activates/deactivates map plotting 
| `.saveInitialTime`     | defines when saving starts 
| `.saveInterval`        | defines saving frequency
Simulation        |                        |
| `seedingAlgorithm`     | dispersal type (see above)
| `successionTimestep`   | defines frequency of dispersal/local recruitment event (growth and mortality are always yearly)
Other             |                        |
| `mixedType`            | how mixed forest stands are defined
| `vegLeadingProportion` | relative biomass threshold to consider a species "leading" (i.e. dominant)


```{r moduleParams, echo = FALSE, eval = TRUE}
df_params <- SpaDES.core::moduleParams("Biomass_core", "..")
knitr::kable(df_params)
```

## Initialization

Unlike the initialization in LBSE, which "iterates the number of time steps equal to the maximum cohort age for each site", beginning at t - oldest cohort age and adding cohorts at the appropriate time ([Scheller & Miranda 2016](https://studylib.net/doc/6761603/landis-ii-biomass-succession-v3.2-user-guide)), Biomass_core initializes the simulation by deriving initial biomasses from available data.

## Compartmentalisation and modularity
Unlike in LBSE, post-disturbance regeneration is not part of Biomass_core *per se*, but belongs to a separate module ([Biomass_regeneration](https://github.com/PredictiveEcology/Biomass_regeneration)), that needs to be loaded should the user want to simulate disturbance effect (*i.e.*, fire disturbances). Again, this enables higher flexibility when swapping between different approaches to regeneration.
For instance, default (*i.e.* not climate sensitive) growth and mortality functions are part of the `LandR` R package, which needs to be loaded prior to running `Biomass_core`. Should the user wish to change the growth/mortality algorithms, they would need to provide compatible functions (with the same names) to the simulation via `simInit` - user-provided functions will replace those loaded with a package.
+ Note: The `LandR` package provides other supporting functions and objects to the simulation, and still needs to be loaded prior to running `Biomass_core`.

## Algorithm improvements
Upon porting LBSE into R, we made five minor modifications to the original model algorithms to better reflect ecological processes. This did not result in dramatic changes in simulation outputs. Briefly, differences with regards to LBSE are:
1. in LandR Biomass growth, mortality and the competition index are calculated at the same time across all cohorts and species, rather than following an age-descending order and input species list order;
2. new cohorts are never included in the calculation of total biomass (*sumB*) when assigning biomass to (other) new cohorts;
3. serotiny and resprouting can occur in the same pixel, albeit for different species only (note that these were moved to another module).
4. shade tolerance values (`sufficientLight` table) can take decimal values for fine tuning.
5. hashing was introduced to reduce over head - pixels are grouped if they share the same cohort structure (i.e. species, age and biomass composition) and the same ecolocation
6. caching of operations related to preparation of inputs (mostly in other modules, but also when sourcing default data in `Biomass_core`)


# Simulation flow

## No disturbances 

`Biomass_core` itself does not simulate disturbances, or their effect on vegetation (*i.e.* post-disturbance mortality and regeneration). The general flow of `Biomass_core` processes is:

1. Preparation of necessary objects for the simulation - either by accessory data prep. modules, or Biomass_core itself (using LANDIS-II test parameters and dummy data for stand age, biomass and land cover and ecological zoning)
2. Seed dispersal - see Section 4.5.1 Seeding of the LANDIS-II Model v7.0 Description(https://drive.google.com/file/d/15gSueug-Rj9I2RZqdroDbad-k53Jq7j3/view) for details
+ Seed dispersal can be a slow process and has been adapted to occur every 10 years. The user can set it to occur more often, but this should not make much of a difference to model outputs, because age classes are meant to be collapsed to tens.
3. Growth and mortality - based on [Scheller and Mladenoff (2004)](https://pdfs.semanticscholar.org/4d38/d0be6b292eccd444af399775d37a757d1967.pdf)
+ unlike dispersal, growth and mortality should occur every year
4. Ageing - based on [Scheller and Mladenoff (2004)](https://pdfs.semanticscholar.org/4d38/d0be6b292eccd444af399775d37a757d1967.pdf)
+ follows the same frequency as dispersal, collapsing ages to classes with resolution = to this frequency
5. Preparation of visual/saved outputs
... (repeat 2-4) ...

## With disturbances 

Note that should a post-disturbance regeneration module be used, regeneration will occur after the disturbance, but *before* dispersal and background vegetation growth and mortality. Hence, the disturbance should take place either at the very beginning or at the very end of each simulation time step. The general flow of Biomass_core processes when disturbances are included (by linking other modules) is:

1. Preparation of necessary objects for the simulation - either by accessory prep. data modules, or `Biomass_core` itself (using LANDIS-II test parameters and dummy data.)
2. Disturbances - simulated by a disturbance module
3. Post-disturbance regeneration - simulated by a regeneration module (`Biomass_regeneration` is an optional download)
4. Seed dispersal - see Section 4.5.1 Seeding of the LANDIS-II Model v7.0 Description(https://drive.google.com/file/d/15gSueug-Rj9I2RZqdroDbad-k53Jq7j3/view) for details
5. Growth, ageing and mortality - based on [Scheller and Mladenoff (2004)](https://pdfs.semanticscholar.org/4d38/d0be6b292eccd444af399775d37a757d1967.pdf)
6. Preparation of visual/saved outputs
... (repeat 2-6) ...

# Usage example

## Load SpaDES

```{r load-SpaDES, eval=FALSE}
library(SpaDES)

moduleName <- c("Biomass_core")  
spadesModulesDirectory <- ".." # In general, a module code will be controlled at one level above the source code
```

## Get the module

See [SpaDES-modules repository](https://github.com/PredictiveEcology/SpaDES-modules) to see how to download this and other SpaDES modules. Alternatively, it can be forked or cloned from github.com directly.

## Setup simulation

```{r module usage example, eval=FALSE}
tempDir <- tempdir()
setPaths(inputPath = file.path(tempDir, "inputs"), 
         cachePath = file.path(tempDir, "cache"), 
         modulePath = spadesModulesDirectory, 
         outputPath = file.path(tempDir, "outputs"))

times <- list(start = 0, end = 10)

studyArea <- Cache(randomStudyArea, size = 1e7) # cache this so it creates a random one only once on a machine

# Pick the species you want to work with -- using the naming convention in "Boreal" column of LandR::sppEquivalencies_CA
speciesNameConvention <- "Boreal"
speciesToUse <- c("Pice_Gla", "Popu_Tre", "Pinu_Con")

sppEquiv <- LandR::sppEquivalencies_CA[get(speciesNameConvention) %in% speciesToUse]
# Assign a colour convention for graphics for each species
sppColorVect <- LandR::sppColors(sppEquiv, speciesNameConvention,
                                 newVals = "Mixed", palette = "Set1")

## Usage example
modules <- as.list(moduleName)
objects <- list(studyArea = studyArea, sppEquiv = sppEquiv, sppColorVect = sppColorVect)
paths <- getPaths()

successionTimestep <- 20L

## keep default values for most parameters 
## (ommitted from this list)
parameters <- list(
  Biomass_core = list(
    "sppEquivCol" = speciesNameConvention
    , "successionTimestep" = successionTimestep
    , ".plots" = c("screen", "object")
    , ".plotInitialTime" = times$start
    , ".plotInterval" = 1L
    , ".saveInitialTime" = times$start
    , ".useCache" = "init"
    , ".useParallel" = FALSE
  )
)

outputs <- data.frame(expand.grid(objectName = "cohortData",
                                  saveTime = unique(seq(times$start, times$end, by = 1)),
                                  eventPriority = 1,
                                  stringsAsFactors = FALSE))

graphics.off()
mySim <- simInitAndSpades(times = times,
                          params = parameters, 
                          modules = modules, 
                          objects = objects, 
                          paths = paths,
                          outputs = outputs,
                          debug = TRUE)
```

# Events
Events are scheduled as follows:

- Module initiation
- Seed dispersal (every `successionTimestep`)
- Mortality and growth
- Reclassification of age cohorts (every `successionTimestep`)
- Summary tables of regeneration (summaryRegen), biomass, age, growth and mortality (summaryBGM*)
- Plots of maps (plotMaps) and averages (plotAvgs)
- Save

# Outputs

The module produces the following outputs:
- Plotting - live and/or saved plot objects/images (depending on `.plots`)
- Saved biomass, mortality, leading vegetation raster layers
- Whatever objects supplied to `outputs` argument in `simInit` (only `cohortData` in the example above)

```{r moduleOutputs, echo = FALSE, eval = TRUE}
df_outputs <- SpaDES.core::moduleOutputs("Biomass_core", "..")
knitr::kable(df_outputs)
```

# Links to other modules

Intended to be used with other landscape modules, such as `LandMine`, `FireSense`, `Biomass_borealDataPrep`, `Biomass_regeneration` and possibly many others.


## Getting help

- <https://gitter.im/PredictiveEcology/LandR_Biomass>

