---
title: "LBMR"
author: "Jean Marchal (jean.d.marchal@gmail.com)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Overview

A biomass succession model derived and modified from LANDIS-II.

# Download the module

```{r download module, eval = FALSE, echo = TRUE}
library(SpaDES)
moduleName <- "LBMR"
spadesModulesDirectory <- "modules" # where the module will be downloaded

# Simple way to download module
downloadModule(moduleName, path = spadesModulesDirectory)

# More advanced use is to fork the github repository, clone, and load.
# The most powerful way is to make a new, empty project of your own name in github,
#  Make a submfolder called "modules" (same as the above name), 
#  then add this module as a submodule inside "modules". This module 
#  then becomes one of many inside your project

```

# Usage

## Module parameters

Name                      | Default           | Description
--------------------------|-------------------|-------------------------------------------
`growthInitialTime`       | `0`               | initial time for the growth event to occur.
`.plotInitialTime`        | `0`               | describes the simulation time at which the first plot event should occur.
`.saveInitialTime`        | `0`               | describes the simulation time at which the first save event should occur.
`spinupMortalityfraction` | `0.001`           | defines the mortality loss fraction in spin up-stage simulation.
`successionTimestep`      | `10`              | defines the simulation time step.
`cellSize`                | `NA`              | defines the cell size.
`seedingAlgorithm`        | `"wardDispersal"` | chooses which seeding algorithm will be used among `"noDispersal"`, `"universalDispersal"`, and `"wardDispersal"`.
`useCache`                | `TRUE`            | use caching for the spinup simulation?
`useParallel`             | `TRUE`            | determines whether the parallel computation will be used in the simulation.

## Usage example

```{r module usage example, eval = FALSE}

inputDir <- file.path(dirname(spadesModulesDirectory), "inputs") %>% checkPath(create = TRUE)
outputDir <- file.path(dirname(spadesModulesDirectory), "outputs") 
times <- list(start = 0, end = 10)
parameters <- list(
  .globals = list(verbose = FALSE),
  LBMR = list(.plotInitialTime = 0,
              successionTimestep = 5,
              seedingAlgorithm = "wardDispersal")
  #.progress = list(type = "text", interval = 1), # for a progress bar
  ## If there are further modules, each can have its own set of parameters:
  #module1 = list(param1 = value1, param2 = value2),
  #module2 = list(param1 = value1, param2 = value2)
)
modules <- list(moduleName)
objects <- list()

setPaths(cachePath = file.path(outputDir, "cache"),
         modulePath = spadesModulesDirectory,
         inputPath = inputDir,
         outputPath = outputDir)
paths <- getPaths()

mySim <- simInit(times = times, params = parameters, modules = modules, objects = objects, paths = paths)

dev()
mySimOut <- spades(mySim, debug = TRUE)
```

# Events
Events are scheduled as follows:

- Module initiation and spin-up
- Account for fire disturbance if present
- Seed dispersal
- Mortality and growth
- Reclassification of age cohorts
- SummaryRegen
- SummaryBGM
- Plot
- Save

# Links to other modules

Intended to be used with other landscape modules, such as LandMine, FireSense, Boreal_LBMRDataPrep, and possibly many others.
